---
title: "Introduction to finitization"
author: "Bogdan Oancea"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to finitization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What is a finitized distribution?

The `finitization` package introduces finitized versions of classical discrete probability distributions.
These versions preserve a given number of moments (e.g., the first 3–5) while allowing for faster random variate generation.

## Basic usage

Load the package:

```{r}
library(finitization)
```

Generate samples from a finitized binomial distribution:

```{r}
set.seed(123)
sample <- rfinitized_binom(n = 5, p = 0.4, order = 3, size = 1000)
hist(sample, breaks = 20, main = "Finitized Binomial Sample", col = "lightblue")
```

Compare it with the standard binomial distribution:

```{r}
standard <- rbinom(1000, size = 5, prob = 0.4)
hist(standard, breaks = 20, main = "Standard Binomial Sample", col = "salmon")
```

You can use similar functions for other distributions:
- `rfinitized_pois()`
- `rfinitized_negbinom()`
- `rfinitized_log()`

## Summary statistics

```{r}
mean(sample)
var(sample)
```

## Visual comparison

```{r}
plot(density(standard), col = "red", lwd = 2, main = "Density Comparison")
lines(density(sample), col = "blue", lwd = 2)
legend("topright", legend = c("Standard", "Finitized"), col = c("red", "blue"), lwd = 2)
```

## Conclusion

Finitized distributions provide a practical approach to generating random variates that approximate key distributional properties efficiently. They are particularly useful in simulation scenarios where performance matters and exact fidelity to all moments is not critical.


# Finitization of Power Series Family of Probability Distributions

## The finitization concept

## Properties of finitized probability distributions

## Usage of finitized probability distributions

## Finitized Poisson distribution

## Finitized Negative Binomial distribution

## Finitized Binomial distribution

## Finitized Logarithmic distribution


